---
title: "DSP Analysis QC Report"
format:
  html:
    code-fold: true
editor: visual

params:
  data.folder: "/Users/cauleyes/CPTR/CPTR-5 Krug_Ramaswami/CPTR_5_DSP_Krug/"
  results.folder: "/Users/cauleyes/CPTR/CPTR-5 Krug_Ramaswami/CPTR_5_DSP_Krug/"
  run.folder: "7_10_2024/"
---

## Load Libraries

```{r Load Libraries}
#| warning: false
#| message: false

# Load all relevant libraries

library(DSPWorkflow)
library(GeomxTools)
library(dplyr)
library(limma)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(ggforce)
library(shadowtext)
library(stringr)
library(PCAtools)
library(readxl)
library(gridExtra)
library(grid)
library(knitr)
library(gt)
library(tidyr)
library(openxlsx)
library(UpSetR)

source("DSP_QC_functions.R")
```

## Initialization

```{r Initialization}

# Input file parameters

pkc.file.name <- "Hs_R_NGS_WTA_v1.0.pkc"
pkc.file.path <- paste0(params$data.folder, pkc.file.name)

annotation.file.name <- "annotation_Krug_CPTR_5_april2024_NC_edit.xlsx"
annotation.file.path <- paste0(params$data.folder, annotation.file.name)

dcc.files <- list.files(file.path(paste0(params$data.folder, "dcc")),
  pattern = ".dcc$",
  full.names = TRUE,
  recursive = TRUE
)

# Annotation parameters
annotation.sheet.name <- "annotation"
sample.id.field.name <- "Sample_ID"
roi.field.name <- "ROI"
panel.field.name <- "panel"
slide.field.name <- "slide name"
class.field.name <- "class"
region.field.name <- "Region"
segment.field.name <- "segment"
area.field.name <- "area"
nuclei.field.name <- "nuclei"
exclude.sankey <- FALSE
segment.id.length <- 8

# Create the GeoMxSet Object

init.object <- initialize_object(dcc.files = dcc.files,
                            pkc.files = pkc.file.path,
                            annotation.file = annotation.file.path,
                            annotation.sheet.name = annotation.sheet.name,
                            sample.id.field.name = sample.id.field.name,
                            roi.field.name = roi.field.name,
                            panel.field.name = panel.field.name,
                            slide.field.name = slide.field.name, 
                            class.field.name = class.field.name, 
                            region.field.name = region.field.name, 
                            segment.field.name = segment.field.name,
                            area.field.name = area.field.name,
                            nuclei.field.name = nuclei.field.name, 
                            segment.id.length = segment.id.length)

```

## Object Summary

@fig-sankey shows a summary of AOIs per annotation

```{r Object Summary}
#| label: fig-sankey
#| fig-cap: "Sankey Plot"
#| warning: false

# Print the Sankey Plot
#sdesign.list$sankey.plot

#Rename the slide name column for formatting
pData(init.object) <- pData(init.object) %>% 
  mutate(slide = gsub("slide_", "", slide_name))

# Define the lanes of the Sankey plot
lane1 <- "slide"
lane2 <- "class"
lane3 <- "region"
lane4 <- "segment"
fill_lane <- "region"

lanes <- c(lane1, lane2, lane3, lane4)


#Establish variables for the Sankey plot
x <- id <- y <- n <- NULL

# select the annotations we want to show, use `` to surround column
# names with spaces or special symbols

# Create a count matrix
count.mat <- count(pData(init.object), 
                   !!as.name(lane1), 
                   !!as.name(lane2), 
                   !!as.name(lane3), 
                   !!as.name(lane4))

# Remove any rows with NA values
na.per.column <- colSums(is.na(count.mat))
na.total.count <- sum(na.per.column)
                                             
if(na.total.count > 0){
  count.mat <- count.mat[!rowSums(is.na(count.mat)),]
  rownames(count.mat) <- 1:nrow(count.mat)
}


# Gather the data and plot in order: lane 1, lane 2, ..., lane n
# gather_set_data creates x, id, y, and n fields within sankey.count.data
# Establish the levels of the Sankey
sankey.count.data <- gather_set_data(count.mat, 1:4)

# Define the annotations to use for the Sankey x axis labels
sankey.count.data$x[sankey.count.data$x == 1] <- lane1
sankey.count.data$x[sankey.count.data$x == 2] <- lane2
sankey.count.data$x[sankey.count.data$x == 3] <- lane3
sankey.count.data$x[sankey.count.data$x == 4] <- lane4

sankey.count.data$x <-
    factor(
      sankey.count.data$x,
      levels = c(as.name(lane1), 
                 as.name(lane2), 
                 as.name(lane3), 
                 as.name(lane4)))
    
# For position of Sankey 100 segment scale
adjust.scale.pos = -1.1

# plot Sankey diagram
sankey.plot <-
  ggplot(sankey.count.data,
         aes(
           x,
           id = id,
           split = y,
           value = n
         )) +
  geom_parallel_sets(aes(fill = !!as.name(fill_lane)), alpha = 0.5, axis.width = 0.1) +
  geom_parallel_sets_axes(axis.width = 0.2, 
                          fill = "seashell", 
                          color = "seashell4") +
  geom_parallel_sets_labels(color = "black",
                            size = 3,
                            angle = 0) + 
  theme_classic(base_size = 14) +
  theme(
    legend.position = "bottom",
    axis.ticks.y = element_blank(),
    axis.line = element_blank(),
    axis.text.y = element_blank()
  ) + 
  scale_y_continuous(expand = expansion(0)) +
  scale_x_discrete(expand = expansion(0)) +
  labs(x = "", y = "") +
  annotate(
    geom = "segment",
    x = (3.25 - adjust.scale.pos),
    xend = (3.25 - adjust.scale.pos),
    y = 20,
    yend = 120,
    lwd = 2
  ) +
  annotate(
    geom = "text",
    x = (3.19 - adjust.scale.pos),
    y = 70,
    angle = 90,
    size = 5,
    hjust = 0.5,
    label = "100 AOIs"
  )

print(sankey.plot)


# Create UpSetR plot for counts per annotation group

AOI.counts <- sankey.count.data

AOI.counts$AOI_count <- AOI.counts$n
AOI.counts$type <- AOI.counts$x
AOI.counts$annotation <- AOI.counts$y

AOI.counts  <-  AOI.counts %>% 
  select(all_of(c("AOI_count", "type", "annotation")))

AOI.count.plot <- ggplot(AOI.counts, aes(x = annotation, y = AOI_count)) + 
  geom_bar(stat = "identity") + 
  facet_wrap(~ type, ncol = 2, scales = "free_x") + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

print(AOI.count.plot)

AOI.inter.counts  <-  sankey.count.data %>% 
  select(all_of(c(lanes, "n")))

AOI.inter.counts <- AOI.inter.counts %>%
  unite("annotation", slide, class, region, segment, sep = "&", remove = FALSE)

# Sum the counts by unique combinations
combined.AOI.inter.counts <- AOI.inter.counts %>%
  group_by(annotation) %>%
  summarize(total_n = sum(n)) %>%
  ungroup()

# Create the named vector
expression.input <- setNames(combined.AOI.inter.counts$total_n, 
                            combined.AOI.inter.counts$annotation)
                            
        

AOI.inter.count.plot <- upset(fromExpression(expression.input),  
                        nsets = 20, 
                        number.angles = 60, 
                        order.by = "freq")

print(AOI.inter.count.plot)

```

## QC and Filtering

```{r QC and Filtering}


qc.output <-  qcProc(object = init.object,
                        min.segment.reads = 1000, 
                        percent.trimmed = 80,    
                        percent.stitched = 80,   
                        percent.aligned = 80,    
                        percent.saturation = 50, 
                        min.negative.count = 3,   
                        max.ntc.count = 1000,     
                        min.nuclei = 200,         
                        min.area = 1000,
                        print.plots = TRUE)
    

```

Summary of QC for AOIs and Probes

```{r QC Summary}

qc.output$table

```

#### AOI QC

AOI distribution by parameter and annotation

```{r AOI Plots}

# Print AOI plots
qc.output$plot$trimmed
qc.output$plot$aligned
qc.output$plot$stitched
qc.output$plot$saturated
qc.output$plot$neg.plot


```

AOIs that have been flagged with the given QC parameters

```{r AOI Flags}

# Print AOI flags

flag.column.detect <- sapply(qc.output$segment.flags, is.logical)
flag.column.names <- names(qc.output$segment.flags[flag.column.detect])

# A function for coloring TRUE flags as red
red.flag <- function(x) {
  x <- as.logical(x)
  ifelse(x, "red", "white") 
  }

# Create the table using the flag coloring function
qc.output$segment.flags %>% 
  gt() %>% 
  data_color(columns = flag.column.names, 
                 fn = red.flag, 
                 alpha = 0.7)

```

### Probe QC

Probes that have been flagged as either local or global outliers.

```{r}

# Create the table for probe flags
probe.flags.df <- qc.output$probe.flags %>% separate_rows(LocalFlag, sep = ",")

# Rename the dcc file name column
probe.flags.df$Sample_ID <- probe.flags.df$LocalFlag

# Grab the annotation for only the columns to map
annotation <- pData(qc.output$object)
annotation$Sample_ID <- rownames(annotation)

annotation.subset <- annotation %>% 
  select(Sample_ID, segmentID)

# Map the AOI names in the flags to the segmentID
probe.flags.df <- merge(probe.flags.df, annotation.subset, by = "Sample_ID")

# Remove the dcc file name column 
probe.flags.table <- probe.flags.df %>% 
  select(TargetName, RTS_ID, segmentID, FlagType)

# For a summary of only probe names
probe.flag.summary <- qc.output$probe.flags %>% 
  select(TargetName, RTS_ID, FlagType)


# Toggle to include all flags or only summary
include.all <- FALSE

# For all flags including segment ID name
if(include.all == TRUE){ 
  
  probe.flags.table %>% 
  gt()
  
} else {
  
  probe.flag.summary %>% 
  gt()
  
}


```

### Filtering

```{r Filtering}

object <- qc.output$object

# Set up lists of segment IDs
segment.list.total <- pData(object)$segmentID

# Define Modules
modules <- gsub(".pkc", "", pkc.file.name)

# Calculate limit of quantification (LOQ) in each segment
# LOQ = geomean(NegProbes) * geoSD(NegProbes)^(LOQ cutoff)
# LOQ is calculated for each module (pkc file)
loq <- data.frame(row.names = colnames(object))

loq.min <- 2
loq.cutoff <- 2

for(module in modules) {
  vars <- paste0(c("NegGeoMean_", "NegGeoSD_"),
                 module)
  if(all(vars[1:2] %in% colnames(pData(object)))) {
    
    neg.geo.mean <- vars[1]
    neg.geo.sd <- vars[2]
    
    loq[, module] <-
      pmax(loq.min,
           pData(object)[, neg.geo.mean] * 
             pData(object)[, neg.geo.sd] ^ loq.cutoff)
  }
}

# Store the loq df in the annotation df
pData(object)$loq <- loq

# Setup a master loq matrix
loq.mat <- c()


for(module in modules) {
  # Gather rows with the given module
  ind <- fData(object)$Module == module
  
  # Check if each feature has counts above the LOQ
  mat.i <- t(esApply(object[ind, ], MARGIN = 1,
                     FUN = function(x) {
                       x > loq[, module]
                     }))
  
  # Store results in the master loq matrix
  loq.mat <- rbind(loq.mat, mat.i)
}

# ensure ordering since this is stored outside of the geomxSet
loq.mat <- loq.mat[fData(object)$TargetName, ]

# Evaluate and Filter Segment Gene Detection Rate
# Save detection rate information to pheno data
pData(object)$GenesDetected <- colSums(loq.mat, na.rm = TRUE)
pData(object)$GeneDetectionRate <- 100*(pData(object)$GenesDetected / nrow(object))

# Establish detection bins
detection.bins <- c("<1", "1-5", "5-10", "10-15", ">15")

# Determine detection thresholds: 1%, 5%, 10%, 15%, >15%
pData(object)$DetectionThreshold <- 
  cut(pData(object)$GeneDetectionRate,
      breaks = c(0, 1, 5, 10, 15, 100),
      labels = detection.bins)

```

#### Overall Gene Detection per AOI

@fig-GeneDetectionbByAOI shows detection rate per AOI, colored by region.

```{r Overall Gene Detection per AOI}
#| label: fig-GeneDetectionbByAOI
#| tbl-cap: "Overall Gene Detection per AOI"
#| warning: false

# stacked bar plot of different cut points (1%, 5%, 10%, 15%)
segment.stacked.bar.plot <- ggplot(pData(object),
                          aes(x = DetectionThreshold)) +
  geom_bar(aes(fill = region)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  theme_bw() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Gene Detection Rate",
         y = "Segments, #",
         fill = "AOI Annotation")


print(segment.stacked.bar.plot)
```

AOIs in the low detection bin of 1-5%

```{r Low Detection AOI}

# cut percent genes detected at 1, 5, 10, 15
segment.table <- kable(table(pData(object)$DetectionThreshold, 
                             pData(object)$class))

# Make a list of segments with low detection
low.detection.segments <- pData(object) %>% 
  filter(GeneDetectionRate < 5) %>% 
  select(any_of(c("segmentID", "GeneDetectionRate")))
rownames(low.detection.segments) <- NULL

# Print low detection segment table
low.detection.segments %>% 
  gt()

```

Gene detection for all AOIs

```{r Gene Detection All AOIs}

# Export a summary of the segment gene detection
segment.detection.summary <- pData(object) %>% 
  select(any_of(c("segmentID", "GeneDetectionRate", "DetectionThreshold")))

```

##### Filter out AOIs with low detection

```{r Filter by AOI}

# Filter the data using the cutoff for gene detection rate
segment.gene.rate.cutoff <- 0

object.segment.filtered <-
    object[, pData(object)$GeneDetectionRate >= segment.gene.rate.cutoff]

```

#### Detection per Gene

```{r Detection per Gene}

# Evaluate and Filter Study-wide Gene Detection Rate 
# Calculate detection rate:
loq.mat <- loq.mat[, colnames(object.segment.filtered)]

fData(object.segment.filtered)$DetectedSegments <- rowSums(loq.mat, na.rm = TRUE)
fData(object.segment.filtered)$DetectionRate <-
  100*(fData(object.segment.filtered)$DetectedSegments / nrow(pData(object)))

# Establish detection bins
detection.bins <- c("0", "<1", "1-5", "5-10", "10-20", "20-30", "30-40", "40-50", ">50")

# Determine detection thresholds: 1%, 5%, 10%, 15%, >15%
fData(object.segment.filtered)$DetectionThreshold <- 
  cut(fData(object.segment.filtered)$DetectionRate,
      breaks = c(-1, 0, 1, 5, 10, 20, 30, 40, 50, 100),
      labels = detection.bins)

```

@fig-DetectionPerGene shows the percent of all AOIs individual genes are detected within

```{r Detection per Gene Plot}
#| label: fig-DetectionPerGene
#| fig-cap: "Gene Detection Percent of All AOIs"
#| warning: false

gene.stacked.bar.plot <- ggplot(fData(object.segment.filtered),
                          aes(x = DetectionThreshold)) +
  geom_bar(aes(fill = Module)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  theme_bw() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Gene Detection Rate",
         y = "Genes, #",
         fill = "Probe Set")

print(gene.stacked.bar.plot)
```

Gene detection rates for specified genes of interest

```{r Genes of Interest}
#| label: tbl-GenesOfInterest
#| tbl-cap: "Gene of Interest Detection Rate"
#| warning: false

# Gene of interest detection table
goi <- c("A2M", "CD44")

goi.table <- data.frame(Gene = goi,
                        Number = fData(object.segment.filtered)[goi, "DetectedSegments"],
                        DetectionRate = fData(object.segment.filtered)[goi, "DetectionRate"])


# Print the GOI table
goi.table %>% 
  gt()
```

@fig-DetectionPerGeneLoss shows the loss of percentage of all AOIs individual genes are detected within

```{r Gene Detection Rates Loss Plot}
#| label: fig-DetectionPerGeneLoss
#| fig-cap: "Gene Detection Percent of All AOIs"
#| warning: false

# Plot detection rate:
plot.detect <- data.frame(Freq = c(1, 5, 10, 20, 30, 50))
plot.detect$Number <-
  unlist(lapply(c(1, 5, 10, 20, 30, 50),
                function(x) {sum(fData(object.segment.filtered)$DetectionRate >= x)}))

plot.detect$Rate <- plot.detect$Number / nrow(fData(object.segment.filtered))
rownames(plot.detect) <- plot.detect$Freq

genes.detected.plot <- ggplot(plot.detect, aes(x = as.factor(Freq), y = Rate, fill = Rate)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = formatC(Number, format = "d", big.mark = ",")),
            vjust = 1.6, color = "black", size = 4) +
  scale_fill_gradient2(low = "orange2", mid = "lightblue",
                       high = "dodgerblue3", midpoint = 0.65,
                       limits = c(0,1),
                       labels = scales::percent) +
  theme_bw() +
  scale_y_continuous(labels = scales::percent, limits = c(0,1),
                     expand = expansion(mult = c(0, 0))) +
  labs(x = "% of Segments",
       y = "Genes Detected, % of Panel > loq")

print(genes.detected.plot)


```

Summary of all gene detection rates

```{r Gene Detection Rate Summary}

# Gather a summary of the every gene's detection percentage in all AOIs
gene.detection.summary <- fData(object.segment.filtered) %>% 
  mutate(gene = rownames(fData(object.segment.filtered))) %>% 
  select(any_of(c("gene", "DetectionRate", "DetectionThreshold")))


```

##### Filter out genes with low detection

```{r Filter Genes}

# Set the cutoff for gene detection
study.gene.rate.cutoff <- 0.00

# Subset for genes above the study gene detection rate cutoff
# Manually include the negative control probe, for downstream use
negative.probe.fData <- subset(fData(object.segment.filtered), CodeClass == "Negative")
neg.probes <- unique(negative.probe.fData$TargetName)
object.gene.filtered <- object.segment.filtered[fData(object.segment.filtered)$DetectionRate >= study.gene.rate.cutoff |
                   fData(object.segment.filtered)$TargetName %in% neg.probes, ]
```

**Write QC Output File**

```{r Write QC output}

# Start the QC output excel workbook
qc.info.output <- createWorkbook()

# Add the AOI flag info to the output file
addWorksheet(qc.info.output, "AOI QC Flags")
writeData(qc.info.output, sheet = "AOI QC Flags", qc.output$segment.flags)

# Add the probe flag QC info to the output file
addWorksheet(qc.info.output, "Probe QC Flags")
writeData(qc.info.output, sheet = "Probe QC Flags", probe.flags.table)

# Add the AOI detection QC info
addWorksheet(qc.info.output, "AOI Detection Rate")
writeData(qc.info.output, sheet = "AOI Detection Rate", segment.detection.summary)

# Add the Gene detection QC info to the output file
addWorksheet(qc.info.output, "Gene Detection Rate")
writeData(qc.info.output, sheet = "Gene Detection Rate", gene.detection.summary)

# Save the QC output file
saveWorkbook(qc.info.output, paste0(params$results.folder, params$run.folder, "QC_info.xlsx"), overwrite = TRUE)
```

### Normalization

```{r Normalization, warning=FALSE, message=FALSE}

q3.normalization.output <- geomxNorm(
                                  object = object.gene.filtered, 
                                  norm = "q3")
    
    
neg.normalization.output <- geomxNorm(
                                  object = object.gene.filtered, 
                                  norm = "neg")
    
```

#### Normalized Reads vs. Negative Background

@fig-Q3vsBackground shows the quartile-3 (Q3) normalization counts compared to negative probes

```{r Mulit-Plot for Q3 Normalization}
#| label: fig-Q3vsBackground
#| fig-cap: "Q3 Normalized Counts versus Background"
#| warning: false

q3.normalization.output$multi.plot


```

@fig-NegvsBackground shows the negative normalization counts compared to negative probes

```{r Mulit-Plot for Neg Normalization}
#| label: fig-NegvsBackground
#| fig-cap: "Negative Normalized Counts versus Background"
#| warning: false

neg.normalization.output$multi.plot
```

**Example AOIs**

@fig-RawReadBoxPlots shows the raw read count boxplots for an example 10 AOIs

```{r Raw reads boxplots}
#| label: fig-RawReadBoxPlots
#| fig-cap: "Raw Counts for an Example 10 AOIs "
#| warning: false

q3.normalization.output$boxplot.raw
```

@fig-Q3NormBoxPlots shows the Q3 normalized read count boxplots for an example 10 AOIs

```{r Q3 norm boxplots}
#| label: fig-Q3NormBoxPlots
#| fig-cap: "Q3 Normalized Counts for an Example 10 AOIs "
#| warning: false

q3.normalization.output$boxplot.norm
```

@fig-NegNormBoxPlots shows the negative normalized read count boxplots for an example 10 AOIs

```{r Neg norm boxplots}
#| label: fig-NegNormBoxPlots
#| fig-cap: "Negative Normalized Counts for an Example 10 AOIs "
#| warning: false

neg.normalization.output$boxplot.norm
```

#### Principal Component Analysis (PCA)

```{r PCA, warning=FALSE, message=FALSE}

# See reference vignette: https://bioconductor.org/packages/release/bioc/vignettes/PCAtools/inst/doc/PCAtools.html#introduction

# Load the Geomx objects
object.q3 <- q3.normalization.output$object
object.neg <- neg.normalization.output$object

# Gather the the normalized counts
q3.norm.counts.df <- as.data.frame(object.q3@assayData$q_norm)
neg.norm.counts.df <- as.data.frame(object.neg@assayData$neg_norm)

# Convert counts to log2
q3.log.counts.df <- q3.norm.counts.df %>% 
  mutate_all(~ log2(.)) %>% 
  rename_all(~ gsub("\\.dcc", "", .))
neg.log.counts.df <- neg.norm.counts.df %>% 
  mutate_all(~ log2(.)) %>% 
  rename_all(~ gsub("\\.dcc", "", .))

# Remove the negative controls from the log counts
control.probes <- c("NegProbe-WTX")
q3.log.counts.df <- q3.log.counts.df[!(rownames(q3.log.counts.df) %in% control.probes), ]
neg.log.counts.df <- neg.log.counts.df[!(rownames(neg.log.counts.df) %in% control.probes), ]

# Load the annotation (same for both normalization types)
annotation <- pData(object.q3)

# Remove NTCs
cleaned.annotation.df <- as.data.frame(annotation[annotation$'slide_name' != "No Template Control", ])

# Order of rownames of annotation need to match columns of count data
cleaned.annotation.df <- cleaned.annotation.df[order(rownames(cleaned.annotation.df)), ]

q3.log.counts.df <- q3.log.counts.df[order(colnames(q3.log.counts.df))]
neg.log.counts.df <- neg.log.counts.df[order(colnames(neg.log.counts.df))]

# Remove .dcc from Sample ID row names
cleaned.annotation.df <- cleaned.annotation.df %>% `rownames<-`(sub("\\.dcc", "", rownames(.)))

# Generate a PCA table for all samples for both normalization types
q3.pca.table <- pca(q3.log.counts.df, 
                 metadata = cleaned.annotation.df, 
                 removeVar = 0.1)
neg.pca.table <- pca(neg.log.counts.df, 
                 metadata = cleaned.annotation.df, 
                 removeVar = 0.1)


```

#### PCA by Segment

```{r PCA for Q3 segment}
#| label: fig-PCAsegmentQ3
#| fig-cap: "PCA colored by Segment for Q3 Normalization"
#| warning: false

q3.pca.plot.segment <- biplot(q3.pca.table, 
                         colby = "segment", 
                         legendPosition = "right", 
                         legendLabSize = 6, 
                         legendIconSize = 3, 
                         lab = NULL,
                         title = "Q3 Normalization", 
                         subtitle = "NTCs removed")

print(q3.pca.plot.segment)

```

```{r PCA for Negative segment}
#| label: fig-PCAsegmentNeg
#| fig-cap: "PCA colored by Segment for Negative Normalization"
#| warning: false

neg.pca.plot.segment <- biplot(neg.pca.table, 
                         colby = "segment", 
                         legendPosition = "right", 
                         legendLabSize = 6, 
                         legendIconSize = 3, 
                         lab = NULL,
                         title = "Negative Normalization", 
                         subtitle = "NTCs removed")

print(neg.pca.plot.segment)

```

#### PCA by Region

```{r PCA for Q3 region}
#| label: fig-PCAregionQ3
#| fig-cap: "PCA colored by Region for Q3 Normalization"
#| warning: false

q3.pca.plot.region <- biplot(q3.pca.table, 
                         colby = "region", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Q3 Normalization", 
                         subtitle = "NTCs removed")


print(q3.pca.plot.region)

```

```{r PCA for Neg region}
#| label: fig-PCAregionNeg
#| fig-cap: "PCA colored by Region for Negative Normalization"
#| warning: false

neg.pca.plot.region <- biplot(neg.pca.table, 
                         colby = "region", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Negative Normalization", 
                         subtitle = "NTCs removed")

print(neg.pca.plot.region)

```

#### PCA by Class

```{r PCA for Q3 class}
#| label: fig-PCAclassQ3
#| fig-cap: "PCA colored by Class for Q3 Normalization"
#| warning: false

q3.pca.plot.class <- biplot(q3.pca.table, 
                         colby = "class", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Q3 Normalization", 
                         subtitle = "NTCs removed")

print(q3.pca.plot.class)

```

```{r PCA for Negative class}
#| label: fig-PCAclassNeg
#| fig-cap: "PCA colored by Class for Negative Normalization"
#| warning: false

neg.pca.plot.class <- biplot(neg.pca.table, 
                         colby = "class", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Negative Normalization", 
                         subtitle = "NTCs removed")

print(neg.pca.plot.class)

```

#### PCA by Slide

```{r PCA for Q3 slide}
#| label: fig-PCAslideQ3
#| fig-cap: "PCA colored by Slide for Q3 Normalization"
#| warning: false


q3.pca.plot.slide <- biplot(q3.pca.table, 
                         colby = "slide_name", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Q3 Normalization", 
                         subtitle = "NTCs removed")

print(q3.pca.plot.slide)

```

```{r PCA for Neg slide}
#| label: fig-PCAslideNeg
#| fig-cap: "PCA colored by Slide for Negative Normalization"
#| warning: false

neg.pca.plot.slide <- biplot(neg.pca.table, 
                         colby = "slide_name", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Negative Normalization", 
                         subtitle = "NTCs removed")

print(neg.pca.plot.slide)

```

### MA Plots

```{r}
# Set up pre and post normalization counts and convert to log

pre.norm.counts <-
  log(as.data.frame(q3.normalization.output$object@assayData$exprs), base = 2)

post.q3norm.counts <- log(as.data.frame(q3.normalization.output$object@assayData$q_norm), base = 2)

post.negnorm.counts <- log(as.data.frame(neg.normalization.output$object@assayData$neg_norm), base = 2)

# Setup annotation groups for the ratio "M" (log A - log B)
tumor.types <- c("Biopsy")
region.types <- c("tumor", "vessel")
source.types <- c("Skin")
segment.types <- c("full_ROI")

# Define the name of the contrast
contrast.name <- paste0(tumor.types[[1]], 
                        "_",
                        source.types[[1]], 
                        "_", 
                        segment.types[[1]], 
                        "_", 
                        region.types[[1]], 
                        "_", 
                        region.types[[2]])

# Set up the annotations and raw counts for the MA plots
contrast.field <- "region"
condition.label <- "tumor"
reference.label <- "vessel"
raw.log.counts <- pre.norm.counts
annotation.MA <- annotation
```

#### Q3 Normalization

```{r MA Plot Q3, message=FALSE, warning=FALSE}
# MA plots for two annotation groups to evaluate negative normalization
log.counts <- post.q3norm.counts

MA.plots.q3 <- make_MA(contrast.field = contrast.field, 
                       condition.label = condition.label, 
                       reference.label = reference.label, 
                       log.counts = log.counts, 
                       raw.log.counts = raw.log.counts, 
                       annotation = annotation.MA)
  
  
grid.draw(MA.plots.q3)
```

#### Negative Normalization

grid.draw(MA.plots.neg)

```{r MA Plot Neg, message=FALSE, warning=FALSE}

# MA plots for two annotation groups to evaluate negative normalization
log.counts <- post.negnorm.counts

MA.plots.neg <- make_MA(contrast.field = contrast.field, 
                    condition.label = condition.label, 
                    reference.label = reference.label, 
                    log.counts = log.counts, 
                    raw.log.counts = raw.log.counts, 
                    annotation = annotation.MA)
  
  
grid.draw(MA.plots.neg)
```
