---
title: "DSP Analysis QC Report Template"
format:
  html:
    code-fold: true
editor: visual
#css: "qc_report_style.css"

params:
  data.folder: "$path_to_folder"
  results.folder: "$path_to_folder"
  run.folder: "$path_to_folder"
---

## Load Libraries

```{r Load Libraries}
#| warning: false
#| message: false

# Load all relevant libraries

library(DSPWorkflow)
library(GeomxTools)
library(dplyr)
library(limma)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(ggforce)
library(shadowtext)
library(stringr)
library(PCAtools)s
library(readxl)
library(gridExtra)
library(grid)
library(knitr)
library(gt)
library(tidyr)
library(openxlsx)
library(ComplexUpset)
library(reshape2)
library(cowplot)

source("DSP_QC_functions.R")
```

## Initialization

```{r Initialization}

# Input file parameters

pkc.file.name <- "Hs_R_NGS_WTA_v1.0.pkc"
pkc.file.path <- paste0(params$data.folder, pkc.file.name)

annotation.file.name <- "$annotation.file.xlsx"
annotation.file.path <- paste0(params$data.folder, annotation.file.name)

dcc.files <- list.files(file.path(paste0(params$data.folder, "dcc")),
  pattern = ".dcc$",
  full.names = TRUE,
  recursive = TRUE
)

# Annotation parameters
annotation.sheet.name <- "annotation"
sample.id.field.name <- "Sample_ID"
roi.field.name <- "ROI"
panel.field.name <- "panel"
slide.field.name <- "slide name"
class.field.name <- "class"
region.field.name <- "Region"
segment.field.name <- "segment"
area.field.name <- "area"
nuclei.field.name <- "nuclei"
exclude.sankey <- FALSE
segment.id.length <- 8

# Create the GeoMxSet Object

init.object <- initialize_object(dcc.files = dcc.files,
                            pkc.files = pkc.file.path,
                            annotation.file = annotation.file.path,
                            annotation.sheet.name = annotation.sheet.name,
                            sample.id.field.name = sample.id.field.name,
                            roi.field.name = roi.field.name,
                            panel.field.name = panel.field.name,
                            slide.field.name = slide.field.name, 
                            class.field.name = class.field.name, 
                            region.field.name = region.field.name, 
                            segment.field.name = segment.field.name,
                            area.field.name = area.field.name,
                            nuclei.field.name = nuclei.field.name, 
                            segment.id.length = segment.id.length)

```

## Object Summary

@fig-sankey shows a summary of AOIs per annotation

```{r Object Summary, fig.width=12, fig.height=8}
#| label: fig-sankey
#| fig-cap: "Sankey Plot"
#| warning: false

sankey.output <- plot_sankey(object = init.object, 
                      lane.1 = "slide", 
                      lane.2 = "class", 
                      lane.3 = "region", 
                      lane.4 = "segment", 
                      fill.lane = "region")

print(sankey.output$sankey.plot)
```

@fig-aoibarplot shows the total AOI counts per annotation

```{r AOI Count Bar Plot, fig.width=12, fig.height=8}
#| label: fig-aoibarplot
#| fig-cap: "AOI Count Bar Plot"
#| warning: false

print(sankey.output$AOI.bar.plot)
```

@fig-upsetr shows the size of annotation groups

```{r UpsetR Plot, fig.width=12, fig.height=8}
#| label: fig-upsetr
#| fig-cap: "UpSetR Plot"
#| warning: false

upsetr.plot <- upsetr_plot(object = init.object, 
                           annotation.groups = c("class", 
                                                 "region", 
                                                 "segment"))


```

## QC and Filtering

```{r QC and Filtering}


qc.output <-  qcProc(object = init.object,
                        min.segment.reads = 1000, 
                        percent.trimmed = 80,    
                        percent.stitched = 80,   
                        percent.aligned = 80,    
                        percent.saturation = 50, 
                        min.negative.count = 3,   
                        max.ntc.count = 1000,     
                        min.nuclei = 200,         
                        min.area = 1000,
                        print.plots = TRUE)
    

```

Summary of QC for AOIs and Probes

```{r QC Summary}

qc.output$table

```

#### AOI QC

AOI distribution by parameter and annotation

```{r AOI Plots}

# Print AOI plots
qc.output$plot$trimmed
qc.output$plot$aligned
qc.output$plot$stitched
qc.output$plot$saturated
qc.output$plot$neg.plot


```

AOIs that have been flagged with the given QC parameters

```{r AOI Flags}

# Print AOI flags

flag.column.detect <- sapply(qc.output$segment.flags, is.logical)
flag.column.names <- names(qc.output$segment.flags[flag.column.detect])

# A function for coloring TRUE flags as red
red.flag <- function(x) {
  x <- as.logical(x)
  ifelse(x, "red", "white") 
  }

# Create the table using the flag coloring function
qc.output$segment.flags %>% 
  gt() %>% 
  data_color(columns = flag.column.names, 
                 fn = red.flag, 
                 alpha = 0.7)

```

### Probe QC

Probes that have been flagged as either local or global outliers.

```{r}

# Create the table for probe flags
probe.flags.df <- qc.output$probe.flags %>% separate_rows(LocalFlag, sep = ",")

# Rename the dcc file name column
probe.flags.df$Sample_ID <- probe.flags.df$LocalFlag

# Grab the annotation for only the columns to map
annotation <- pData(qc.output$object)
annotation$Sample_ID <- rownames(annotation)

annotation.subset <- annotation %>% 
  select(Sample_ID, segmentID)

# Map the AOI names in the flags to the segmentID
probe.flags.df <- merge(probe.flags.df, annotation.subset, by = "Sample_ID")

# Remove the dcc file name column 
probe.flags.table <- probe.flags.df %>% 
  select(TargetName, RTS_ID, segmentID, FlagType)

# For a summary of only probe names
probe.flag.summary <- qc.output$probe.flags %>% 
  select(TargetName, RTS_ID, FlagType)


# Toggle to include all flags or only summary
include.all <- FALSE

# For all flags including segment ID name
if(include.all == TRUE){ 
  
  probe.flags.table %>% 
  gt()
  
} else {
  
  probe.flag.summary %>% 
  gt()
  
}


```

### Filtering

```{r Filtering}

object <- qc.output$object

# Set up lists of segment IDs
segment.list.total <- pData(object)$segmentID

# Define Modules
modules <- gsub(".pkc", "", pkc.file.name)

# Calculate limit of quantification (LOQ) in each segment
# LOQ = geomean(NegProbes) * geoSD(NegProbes)^(LOQ cutoff)
# LOQ is calculated for each module (pkc file)
loq <- data.frame(row.names = colnames(object))

loq.min <- 2
loq.cutoff <- 2

for(module in modules) {
  vars <- paste0(c("NegGeoMean_", "NegGeoSD_"),
                 module)
  if(all(vars[1:2] %in% colnames(pData(object)))) {
    
    neg.geo.mean <- vars[1]
    neg.geo.sd <- vars[2]
    
    loq[, module] <-
      pmax(loq.min,
           pData(object)[, neg.geo.mean] * 
             pData(object)[, neg.geo.sd] ^ loq.cutoff)
  }
}

# Store the loq df in the annotation df
pData(object)$loq <- loq

# Setup a master loq matrix
loq.mat <- c()


for(module in modules) {
  # Gather rows with the given module
  ind <- fData(object)$Module == module
  
  # Check if each feature has counts above the LOQ
  mat.i <- t(esApply(object[ind, ], MARGIN = 1,
                     FUN = function(x) {
                       x > loq[, module]
                     }))
  
  # Store results in the master loq matrix
  loq.mat <- rbind(loq.mat, mat.i)
}

# ensure ordering since this is stored outside of the geomxSet
loq.mat <- loq.mat[fData(object)$TargetName, ]

# Evaluate and Filter Segment Gene Detection Rate
# Save detection rate information to pheno data
pData(object)$GenesDetected <- colSums(loq.mat, na.rm = TRUE)
pData(object)$GeneDetectionRate <- 100*(pData(object)$GenesDetected / nrow(object))

# Establish detection bins
detection.bins <- c("<1", "1-5", "5-10", "10-15", ">15")

# Determine detection thresholds: 1%, 5%, 10%, 15%, >15%
pData(object)$DetectionThreshold <- 
  cut(pData(object)$GeneDetectionRate,
      breaks = c(0, 1, 5, 10, 15, 100),
      labels = detection.bins)

```

#### Overall Gene Detection per AOI

@fig-GeneDetectionbByAOI shows detection rate per AOI, colored by region.

```{r Overall Gene Detection per AOI}
#| label: fig-GeneDetectionbByAOI
#| tbl-cap: "Overall Gene Detection per AOI"
#| warning: false

# stacked bar plot of different cut points (1%, 5%, 10%, 15%)
segment.stacked.bar.plot <- ggplot(pData(object),
                          aes(x = DetectionThreshold)) +
  geom_bar(aes(fill = region)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  theme_bw() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Gene Detection Rate",
         y = "Segments, #",
         fill = "AOI Annotation")


print(segment.stacked.bar.plot)
```

AOIs in the low detection bin of 1-5%

```{r Low Detection AOI}

# cut percent genes detected at 1, 5, 10, 15
segment.table <- kable(table(pData(object)$DetectionThreshold, 
                             pData(object)$class))

# Make a list of segments with low detection
low.detection.segments <- pData(object) %>% 
  filter(GeneDetectionRate < 5) %>% 
  select(any_of(c("segmentID", "GeneDetectionRate")))
rownames(low.detection.segments) <- NULL

# Print low detection segment table
low.detection.segments %>% 
  gt()

```

Gene detection for all AOIs

```{r Gene Detection All AOIs}

# Export a summary of the segment gene detection
segment.detection.summary <- pData(object) %>% 
  select(any_of(c("segmentID", "GeneDetectionRate", "DetectionThreshold")))

```

##### Filter out AOIs with low detection

```{r Filter by AOI}

# Filter the data using the cutoff for gene detection rate
segment.gene.rate.cutoff <- 0

object.segment.filtered <-
    object[, pData(object)$GeneDetectionRate >= segment.gene.rate.cutoff]

```

#### Detection per Gene

```{r Detection per Gene}

# Evaluate and Filter Study-wide Gene Detection Rate 
# Calculate detection rate:
loq.mat <- loq.mat[, colnames(object.segment.filtered)]

fData(object.segment.filtered)$DetectedSegments <- rowSums(loq.mat, na.rm = TRUE)
fData(object.segment.filtered)$DetectionRate <-
  100*(fData(object.segment.filtered)$DetectedSegments / nrow(pData(object)))

# Establish detection bins
detection.bins <- c("0", "<1", "1-5", "5-10", "10-20", "20-30", "30-40", "40-50", ">50")

# Determine detection thresholds: 1%, 5%, 10%, 15%, >15%
fData(object.segment.filtered)$DetectionThreshold <- 
  cut(fData(object.segment.filtered)$DetectionRate,
      breaks = c(-1, 0, 1, 5, 10, 20, 30, 40, 50, 100),
      labels = detection.bins)

```

@fig-DetectionPerGene shows the percent of all AOIs individual genes are detected within

```{r Detection per Gene Plot}
#| label: fig-DetectionPerGene
#| fig-cap: "Gene Detection Percent of All AOIs"
#| warning: false

gene.stacked.bar.plot <- ggplot(fData(object.segment.filtered),
                          aes(x = DetectionThreshold)) +
  geom_bar(aes(fill = Module)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  theme_bw() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Gene Detection Rate",
         y = "Genes, #",
         fill = "Probe Set")

print(gene.stacked.bar.plot)
```

Gene detection rates for specified genes of interest

```{r Genes of Interest}
#| label: tbl-GenesOfInterest
#| tbl-cap: "Gene of Interest Detection Rate"
#| warning: false

# Gene of interest detection table
goi <- c("A2M", "CD44")

goi.table <- data.frame(Gene = goi,
                        Number = fData(object.segment.filtered)[goi, "DetectedSegments"],
                        DetectionRate = fData(object.segment.filtered)[goi, "DetectionRate"])


# Print the GOI table
goi.table %>% 
  gt()
```

@fig-DetectionPerGeneLoss shows the loss of percentage of all AOIs individual genes are detected within

```{r Gene Detection Rates Loss Plot}
#| label: fig-DetectionPerGeneLoss
#| fig-cap: "Gene Detection Percent of All AOIs"
#| warning: false

# Plot detection rate:
plot.detect <- data.frame(Freq = c(1, 5, 10, 20, 30, 50))
plot.detect$Number <-
  unlist(lapply(c(1, 5, 10, 20, 30, 50),
                function(x) {sum(fData(object.segment.filtered)$DetectionRate >= x)}))

plot.detect$Rate <- plot.detect$Number / nrow(fData(object.segment.filtered))
rownames(plot.detect) <- plot.detect$Freq

genes.detected.plot <- ggplot(plot.detect, aes(x = as.factor(Freq), y = Rate, fill = Rate)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = formatC(Number, format = "d", big.mark = ",")),
            vjust = 1.6, color = "black", size = 4) +
  scale_fill_gradient2(low = "orange2", mid = "lightblue",
                       high = "dodgerblue3", midpoint = 0.65,
                       limits = c(0,1),
                       labels = scales::percent) +
  theme_bw() +
  scale_y_continuous(labels = scales::percent, limits = c(0,1),
                     expand = expansion(mult = c(0, 0))) +
  labs(x = "% of Segments",
       y = "Genes Detected, % of Panel > loq")

print(genes.detected.plot)


```

Summary of all gene detection rates

```{r Gene Detection Rate Summary}

# Gather a summary of the every gene's detection percentage in all AOIs
gene.detection.summary <- fData(object.segment.filtered) %>% 
  mutate(gene = rownames(fData(object.segment.filtered))) %>% 
  select(any_of(c("gene", "DetectionRate", "DetectionThreshold")))


```

##### Filter out genes with low detection

```{r Filter Genes}

# Set the cutoff for gene detection
study.gene.rate.cutoff <- 0.00

# Subset for genes above the study gene detection rate cutoff
# Manually include the negative control probe, for downstream use
negative.probe.fData <- subset(fData(object.segment.filtered), CodeClass == "Negative")
neg.probes <- unique(negative.probe.fData$TargetName)
object.gene.filtered <- object.segment.filtered[fData(object.segment.filtered)$DetectionRate >= study.gene.rate.cutoff |
                   fData(object.segment.filtered)$TargetName %in% neg.probes, ]
```

**Write QC Output File**

```{r Write QC output}

# Start the QC output excel workbook
qc.info.output <- createWorkbook()

# Add the AOI flag info to the output file
addWorksheet(qc.info.output, "AOI QC Flags")
writeData(qc.info.output, sheet = "AOI QC Flags", qc.output$segment.flags)

# Add the probe flag QC info to the output file
addWorksheet(qc.info.output, "Probe QC Flags")
writeData(qc.info.output, sheet = "Probe QC Flags", probe.flags.table)

# Add the AOI detection QC info
addWorksheet(qc.info.output, "AOI Detection Rate")
writeData(qc.info.output, sheet = "AOI Detection Rate", segment.detection.summary)

# Add the Gene detection QC info to the output file
addWorksheet(qc.info.output, "Gene Detection Rate")
writeData(qc.info.output, sheet = "Gene Detection Rate", gene.detection.summary)

# Save the QC output file
saveWorkbook(qc.info.output, paste0(params$results.folder, params$run.folder, "QC_info.xlsx"), overwrite = TRUE)
```

#### Q3 versus Negative Background

```{r Visualize Density of Q3 Scores versus Negative Background, warning=FALSE, message=FALSE}
#| label: fig-densityQ3vsBackground
#| fig-cap: "Density of Q3 Normalized Counts versus Background"
#| warning: false

# Object to use for counts
object <- object.gene.filtered

# Annotation to use in the plots
facet.annotation <- "region"

# Set up variables for computing stat data
color.variable <- Value <- Statistic <- NegProbe <- Q3 <- Annotation <- NULL
neg.probes<- "NegProbe-WTX"

# Compute the stat data
stat.data <- base::data.frame(row.names = colnames(exprs(object)),
                                AOI = colnames(exprs(object)),
                                Annotation = Biobase::pData(object)[, facet.annotation],
                                Q3 = unlist(apply(exprs(object), 2,
                                                  quantile, 0.75, na.rm = TRUE)),
                                NegProbe = exprs(object)[neg.probes, ])
  
# Melt stat data for easier plotting
stat.data.melt <- melt(stat.data, measures.vars = c("Q3", "NegProbe"),
                      variable.name = "Statistic", value.name = "Value")


# Compute means for each annnotation group and negative background
stat.data.mean <- stat.data.melt %>% 
    mutate(group = paste0(Annotation, Statistic)) %>% 
    group_by(group) %>% 
    mutate(group_mean = mean(Value)) %>% 
    ungroup() %>% 
    select(Annotation, Statistic, group_mean) %>% 
    distinct()

# Plot with annotation groups separated
distribution.plot <- ggplot(stat.data.melt, aes(x=Value, 
                                               color=Statistic, 
                                               fill=Statistic)) + 
    geom_density(alpha=0.6) +
    geom_vline(data=stat.data.mean, aes(xintercept=group_mean, color=Statistic),
               linetype="dashed") +
    scale_color_manual(values = c("#56B4E9", "#E69F00")) +
    scale_fill_manual(values=c("#56B4E9", "#E69F00")) + 
    scale_x_continuous(trans = "log2") +  
    facet_wrap(~Annotation, nrow = 1) + 
    labs(title=paste0("Density of AOI counts Q3 vs Negative by ", facet.annotation), 
         x="Probe Counts per AOI", 
         y = "Density from AOI Count", 
         color = "Statistic", 
         fill = "Statistic") +
    theme_bw()

# Plot overlapping density
distribution.plot.overlap <- ggplot(stat.data.melt, aes(x=Value, 
                                               color=Annotation, 
                                               fill=Annotation)) + 
    geom_density(alpha=0.2) + 
    scale_x_continuous(trans = "log2") + 
    labs(title=paste0("Density of AOI counts Q3 by ", facet.annotation), 
         x="Probe Counts per AOI", 
         y = "Density from AOI Count", 
         color = "Annotation", 
         fill = "Annotation") +
    theme_bw()

# Combine plots into a single output
distr.plots <- plot_grid(distribution.plot, 
                          distribution.plot.overlap, 
                          ncol = 1)

print(distr.plots)

q3.neg.slope.plot <- ggplot(stat.data, 
               aes(x = NegProbe, y = Q3, color = Annotation)) + 
  geom_abline(intercept = 0, 
              slope = 1, 
              lty = "dashed", 
              color = "darkgray") + 
  geom_point() + guides(color = "none") + 
  theme_bw() + 
  scale_x_continuous(trans = "log2") + 
  scale_y_continuous(trans = "log2") + 
  theme(aspect.ratio = 1) + 
  labs(x = "Negative Probe GeoMean, Counts", y = "Q3 Value, Counts")

print(q3.neg.slope.plot)

q3.neg.ratio.plot <- ggplot(stat.data, 
                            aes(x = NegProbe, 
                                y = Q3/NegProbe, 
                                color = Annotation)) + 
  geom_hline(yintercept = 1, 
             lty = "dashed", 
             color = "darkgray") + 
  geom_point() + 
  theme_bw() + 
  scale_x_continuous(trans = "log2") + 
  scale_y_continuous(trans = "log2") + 
  theme(aspect.ratio = 1) + 
  labs(x = "Negative Probe GeoMean, Counts", y = "Q3/NegProbe Value, Counts")

print(q3.neg.ratio.plot)

stat.data <- stat.data %>% 
  mutate(q3_neg_ratio = Q3/NegProbe) %>% 
  mutate(low_ratio_flag = ifelse(q3_neg_ratio < 1.1, 
                                 "TRUE", 
                                 "FALSE"))

```

### Normalization

```{r Normalization, warning=FALSE, message=FALSE}

q3.normalization.output <- geomxNorm(
                                  object = object.gene.filtered, 
                                  norm = "q3")
    
    
neg.normalization.output <- geomxNorm(
                                  object = object.gene.filtered, 
                                  norm = "neg")
    

export.norm.object <- FALSE 
if(export.norm.object == TRUE){
  
  object <- q3.normalization.object
  
  save(object, file = paste0(params$results.folder, params$run.folder, "$normalized.object.RDA"))
  
}   
```

#### **Example AOIs**

```{r Normalization Effects on Counts, fig.width=12, fig.height=8}
#| label: fig-NormEffects
#| fig-cap: "Normalization Effects on Counts"
#| warning: false
#| message: false

# The raw counts boxplot
transform1.raw<- exprs(q3.normalization.object[,1:30])
transform2.raw<- as.data.frame(transform1.raw)
transform3.raw<- melt(transform2.raw)
ggboxplot.raw <- ggplot(transform3.raw, aes(variable, value)) +
  stat_boxplot(geom = "errorbar") +
  geom_boxplot(fill="grey") +
  scale_y_log10() +
  xlab("Example AOIs") + 
  ylab("Counts, Raw") +
  ggtitle("Neg Norm Counts") +
  scale_x_discrete(labels=c(1:30))

# The Q3 normalized counts boxplot
transform1.norm<- assayDataElement(q3.normalization.object[,1:30], elt = "q_norm")
transform2.norm<- as.data.frame(transform1.norm)
transform3.norm<- melt(transform2.norm)
ggboxplot.q3norm <- ggplot(transform3.norm, aes(variable, value)) +
  stat_boxplot(geom = "errorbar") +
  geom_boxplot(fill="cadetblue2") +
  scale_y_log10() +
  xlab("Example AOIs") + 
  ylab("Counts, Q3 Normalized") +
  ggtitle("Q3 Norm Counts") +
  scale_x_discrete(labels=c(1:30))

# The Negative normalized counts boxplot
transform1.norm<- assayDataElement(neg.normalization.object[,1:30], elt = "neg_norm")
transform2.norm<- as.data.frame(transform1.norm)
transform3.norm<- melt(transform2.norm)
ggboxplot.negnorm <- ggplot(transform3.norm, aes(variable, value)) +
  stat_boxplot(geom = "errorbar") +
  geom_boxplot(fill="indianred") +
  scale_y_log10() +
  xlab("Example AOIs") + 
  ylab("Counts, Neg. Normalized") +
  ggtitle("Neg Norm Counts") +
  scale_x_discrete(labels=c(1:30))

print(ggboxplot.raw)
print(ggboxplot.q3norm)
print(ggboxplot.negnorm)


```

#### Principal Component Analysis (PCA)

```{r PCA, warning=FALSE, message=FALSE}

# See reference vignette: https://bioconductor.org/packages/release/bioc/vignettes/PCAtools/inst/doc/PCAtools.html#introduction

# Load the Geomx objects
object.q3 <- q3.normalization.output$object
object.neg <- neg.normalization.output$object

# Gather the the normalized counts
q3.norm.counts.df <- as.data.frame(object.q3@assayData$q_norm)
neg.norm.counts.df <- as.data.frame(object.neg@assayData$neg_norm)

# Convert counts to log2
q3.log.counts.df <- q3.norm.counts.df %>% 
  mutate_all(~ log2(.)) %>% 
  rename_all(~ gsub("\\.dcc", "", .))
neg.log.counts.df <- neg.norm.counts.df %>% 
  mutate_all(~ log2(.)) %>% 
  rename_all(~ gsub("\\.dcc", "", .))

# Remove the negative controls from the log counts
control.probes <- c("NegProbe-WTX")
q3.log.counts.df <- q3.log.counts.df[!(rownames(q3.log.counts.df) %in% control.probes), ]
neg.log.counts.df <- neg.log.counts.df[!(rownames(neg.log.counts.df) %in% control.probes), ]

# Load the annotation (same for both normalization types)
annotation <- pData(object.q3)

# Remove NTCs
cleaned.annotation.df <- as.data.frame(annotation[annotation$'slide_name' != "No Template Control", ])

# Order of rownames of annotation need to match columns of count data
cleaned.annotation.df <- cleaned.annotation.df[order(rownames(cleaned.annotation.df)), ]

q3.log.counts.df <- q3.log.counts.df[order(colnames(q3.log.counts.df))]
neg.log.counts.df <- neg.log.counts.df[order(colnames(neg.log.counts.df))]

# Remove .dcc from Sample ID row names
cleaned.annotation.df <- cleaned.annotation.df %>% `rownames<-`(sub("\\.dcc", "", rownames(.)))

# Generate a PCA table for all samples for both normalization types
q3.pca.table <- pca(q3.log.counts.df, 
                 metadata = cleaned.annotation.df, 
                 removeVar = 0.1)
neg.pca.table <- pca(neg.log.counts.df, 
                 metadata = cleaned.annotation.df, 
                 removeVar = 0.1)


```

#### PCA by Segment

```{r PCA for Q3 segment, fig.width=12, fig.height=8}
#| label: fig-PCAsegmentQ3
#| fig-cap: "PCA colored by Segment for Q3 Normalization"
#| warning: false

q3.pca.plot.segment <- biplot(q3.pca.table, 
                         colby = "segment", 
                         legendPosition = "right", 
                         legendLabSize = 6, 
                         legendIconSize = 3, 
                         lab = NULL,
                         title = "Q3 Normalization", 
                         subtitle = "NTCs removed")

print(q3.pca.plot.segment)

```

```{r PCA for Negative segment, fig.width=12, fig.height=8}
#| label: fig-PCAsegmentNeg
#| fig-cap: "PCA colored by Segment for Negative Normalization"
#| warning: false

neg.pca.plot.segment <- biplot(neg.pca.table, 
                         colby = "segment", 
                         legendPosition = "right", 
                         legendLabSize = 6, 
                         legendIconSize = 3, 
                         lab = NULL,
                         title = "Negative Normalization", 
                         subtitle = "NTCs removed")

print(neg.pca.plot.segment)

```

#### PCA by Region

```{r PCA for Q3 region, fig.width=12, fig.height=8}
#| label: fig-PCAregionQ3
#| fig-cap: "PCA colored by Region for Q3 Normalization"
#| warning: false

q3.pca.plot.region <- biplot(q3.pca.table, 
                         colby = "region", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Q3 Normalization", 
                         subtitle = "NTCs removed")


print(q3.pca.plot.region)

```

```{r PCA for Neg region, fig.width=12, fig.height=8}
#| label: fig-PCAregionNeg
#| fig-cap: "PCA colored by Region for Negative Normalization"
#| warning: false

neg.pca.plot.region <- biplot(neg.pca.table, 
                         colby = "region", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Negative Normalization", 
                         subtitle = "NTCs removed")

print(neg.pca.plot.region)

```

#### PCA by Class

```{r PCA for Q3 class, fig.width=12, fig.height=8}
#| label: fig-PCAclassQ3
#| fig-cap: "PCA colored by Class for Q3 Normalization"
#| warning: false

q3.pca.plot.class <- biplot(q3.pca.table, 
                         colby = "class", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Q3 Normalization", 
                         subtitle = "NTCs removed")

print(q3.pca.plot.class)

```

```{r PCA for Negative class, fig.width=12, fig.height=8}
#| label: fig-PCAclassNeg
#| fig-cap: "PCA colored by Class for Negative Normalization"
#| warning: false

neg.pca.plot.class <- biplot(neg.pca.table, 
                         colby = "class", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Negative Normalization", 
                         subtitle = "NTCs removed")

print(neg.pca.plot.class)

```

#### PCA by Slide

```{r PCA for Q3 slide, fig.width=12, fig.height=8}
#| label: fig-PCAslideQ3
#| fig-cap: "PCA colored by Slide for Q3 Normalization"
#| warning: false


q3.pca.plot.slide <- biplot(q3.pca.table, 
                         colby = "slide_name", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Q3 Normalization", 
                         subtitle = "NTCs removed")

print(q3.pca.plot.slide)

```

```{r PCA for Neg slide, fig.width=12, fig.height=8}
#| label: fig-PCAslideNeg
#| fig-cap: "PCA colored by Slide for Negative Normalization"
#| warning: false

neg.pca.plot.slide <- biplot(neg.pca.table, 
                         colby = "slide_name", 
                         legendPosition = "right", 
                         legendLabSize = 10, 
                         legendIconSize = 5, 
                         lab = NULL,
                         title = "Negative Normalization", 
                         subtitle = "NTCs removed")

print(neg.pca.plot.slide)

```

### MA Plots

```{r}
# Set up pre and post normalization counts and convert to log

pre.norm.counts <-
  log(as.data.frame(q3.normalization.output$object@assayData$exprs), base = 2)

post.q3norm.counts <- log(as.data.frame(q3.normalization.output$object@assayData$q_norm), base = 2)

post.negnorm.counts <- log(as.data.frame(neg.normalization.output$object@assayData$neg_norm), base = 2)

# Setup annotation groups for the ratio "M" (log A - log B)
region.types <- c("region_A", "region_B")

# Define the name of the contrast
contrast.name <- paste0(region.types[[1]], 
                        "_", 
                        region.types[[2]])

# Set up the annotations and raw counts for the MA plots
contrast.field <- "region"
condition.label <- "region_A"
reference.label <- "region_B"
raw.log.counts <- pre.norm.counts
annotation.MA <- annotation
```

#### Q3 Normalization

```{r MA Plot Q3, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
# MA plots for two annotation groups to evaluate negative normalization
log.counts <- post.q3norm.counts

MA.plots.q3 <- make_MA(contrast.field = contrast.field, 
                       condition.label = condition.label, 
                       reference.label = reference.label, 
                       log.counts = log.counts, 
                       raw.log.counts = raw.log.counts, 
                       annotation = annotation.MA)
  
  
grid.draw(MA.plots.q3)
```

#### Negative Normalization

grid.draw(MA.plots.neg)

```{r MA Plot Neg, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}

# MA plots for two annotation groups to evaluate negative normalization
log.counts <- post.negnorm.counts

MA.plots.neg <- make_MA(contrast.field = contrast.field, 
                    condition.label = condition.label, 
                    reference.label = reference.label, 
                    log.counts = log.counts, 
                    raw.log.counts = raw.log.counts, 
                    annotation = annotation.MA)
  
  
grid.draw(MA.plots.neg)
```
