---
title: "DSP DE Analysis - Kidney Test Dataset"
format:
  html:
    code-fold: true
editor: visual

params:
  data.folder: "test_datasets/Human_Kidney/"
  de.folder: "test_datasets/Human_Kidney/de/"
  dataset.name: "human_kidney_test_"
---

# DE Analysis

```{r Setup}
#| warning: false
#| message: false

library(GeomxTools)
library(dplyr)
library(limma)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(stringr)
library(PCAtools)
library(readxl)
library(gridExtra)
library(grid)
library(DSPWorkflow)
library(standR)
library(SpatialExperiment)

# Source the helper functions
source("DSP_functions.R")

#de.results.files <- list.files(params$de.folder)
```

### Load the Normalized Counts and Annotation

```{r Load Normalized Counts}

normalized.object <- readRDS(file = paste0(params$data.folder,
                                                 params$dataset.name, 
                                                 "q3normalized_object.RDS"))


```

## Main DE Analysis Function

```{r de-helper-functions}
#| warning: false
#| message: false

# Use the left-hand value unless it is NULL; otherwise use the default on the right.
`%||%` <- function(x, y) {
  if (is.null(x)) y else x
}

standardize_de_results <- function(results.df) {
  logfc.column <- grep("logFC", colnames(results.df), value = TRUE)
  pval.column <- grep("_pval", colnames(results.df), value = TRUE)
  adj.pval.column <- grep("adjpval", colnames(results.df), value = TRUE)
  
  # Checks
  if (length(logfc.column) != 1) {
    stop("Expected exactly one logFC column, found: ",
         paste(logfc.column, collapse = ", "))
  }

  if (length(pval.column) != 1) {
    stop("Expected exactly one _pval column, found: ",
         paste(pval.column, collapse = ", "))
  }

  if (length(adj.pval.column) != 1) {
    stop("Expected exactly one adjpval column, found: ",
         paste(adj.pval.column, collapse = ", "))
  }

  data.frame(
    gene = results.df$Gene,
    logfc = results.df[[logfc.column]],
    pval = results.df[[pval.column]],
    padj = results.df[[adj.pval.column]],
    stringsAsFactors = FALSE
  )
}

validate_cached_results <- function(results.df, result.file) {
  required.cols <- c("gene", "logfc", "pval", "padj")

  if (!all(required.cols %in% colnames(results.df))) {
    stop(
      "Cached DE result file does not contain the expected columns: ",
      result.file,
      "\nExpected columns: ",
      paste(required.cols, collapse = ", ")
    )
  }

  results.df <- results.df %>%
    dplyr::select(dplyr::all_of(required.cols))

  results.df$logfc <- as.numeric(results.df$logfc)
  results.df$pval <- as.numeric(results.df$pval)
  results.df$padj <- as.numeric(results.df$padj)

  results.df
}

get_top_gene_labels <- function(results.df,
                                fc.limit = 1.5,
                                padj.limit = 0.05,
                                n = 10) {
  down <- results.df %>%
    filter(!is.na(logfc), !is.na(padj)) %>%
    filter(logfc < -fc.limit, padj < padj.limit) %>%
    arrange(padj, logfc)

  up <- results.df %>%
    filter(!is.na(logfc), !is.na(padj)) %>%
    filter(logfc > fc.limit, padj < padj.limit) %>%
    arrange(padj, desc(logfc))

  unique(c(
    head(down$gene, n),
    head(up$gene, n)
  ))
}

make_de_volcano <- function(results.df,
                            comp.params,
                            custom.labels = NULL,
                            remove.all.gene.labels = FALSE,
                            label.size = NULL) {
  improved_make_volcano(
    lmm.results = results.df,
    title = comp.params$volcano.title %||% comp.params$heading %||% comp.params$contrast.name,
    title.size = comp.params$title.size %||% 22,
    legend.title = comp.params$legend.title %||% "Expression",
    fc.limit = comp.params$fc.limit %||% 1.5,
    custom.gene.labels = custom.labels,
    remove.controls = comp.params$remove.controls %||% FALSE,
    remove.genes = comp.params$remove.genes %||% NULL,
    remove.all.gene.labels = remove.all.gene.labels,
    legend.coordinates = comp.params$legend.coordinates %||% c(.99, .1),
    x.lab = comp.params$x.lab %||% "Log2 Fold Change",
    y.lab = comp.params$y.lab %||% "-Log10 adjusted p-value",
    dotted.line.color = comp.params$dotted.line.color %||% "black",
    alpha = comp.params$alpha %||% 1,
    nonDE.color = comp.params$nonDE.color %||% "gray60",
    upDE.color = comp.params$upDE.color %||% "red",
    downDE.color = comp.params$downDE.color %||% "blue",
    label.size = label.size %||% comp.params$label.size %||% 8,
    label.color = comp.params$label.color %||% "black",
    axis.tick.label.size = comp.params$axis.tick.label.size %||% 20,
    legend.text.size = comp.params$legend.text.size %||% 20,
    axis.title.size = comp.params$axis.title.size %||% 20
  )
}

run_de_comparison <- function(comp.params, normalized.object, params) {
  heading <- comp.params$heading %||% comp.params$contrast.name

  cat("\n\n#### ", heading, "\n\n", sep = "")

  lmm.input <- subset_object_for_lmm(
    object = normalized.object,
    subset.group.1 = comp.params$subset.group.1,
    subset.group.2 = comp.params$subset.group.2
  )

  summary.cols <- union(
    names(comp.params$subset.group.1),
    names(comp.params$subset.group.2)
  )

  summary.table.df <- pData(lmm.input$subset.object) %>%
    dplyr::select(dplyr::all_of(summary.cols))

  summary.table <- as.data.frame(
    table(summary.table.df),
    stringsAsFactors = FALSE
  )

  cat("**Sample numbers per annotation group**\n\n")
  cat(knitr::kable(summary.table), sep = "\n")
  cat("\n\n")

  result.file <- file.path(
    params$de.folder,
    paste0(comp.params$contrast.name, "_de.results.csv")
  )

  if (file.exists(result.file) && !isTRUE(comp.params$overwrite.results %||% FALSE)) {
    results.df <- read.csv(
      result.file,
      stringsAsFactors = FALSE,
      check.names = FALSE
    )

    results.df <- validate_cached_results(results.df, result.file)

    cat("*Loaded existing DE results.*\n\n")
  } else {
    results.list <- diffExpr(
      object = lmm.input$subset.object,
      analysis.type = comp.params$analysis.type %||% "Within Groups",
      region.col = comp.params$region.col,
      regions = comp.params$regions,
      group.col = comp.params$group.col,
      groups = comp.params$groups,
      n.cores = comp.params$n.cores %||% parallel::detectCores()
    )

    results.df <- standardize_de_results(results.list$results)

    if (isTRUE(comp.params$write.results %||% TRUE)) {
      dir.create(params$de.folder, showWarnings = FALSE, recursive = TRUE)

      write.csv(
        results.df,
        result.file,
        row.names = FALSE
      )

      cat("*Wrote DE results to disk.*\n\n")
    }
  }

  fc.limit <- comp.params$fc.limit %||% 1.5
  padj.limit <- comp.params$padj.limit %||% 0.05

  total.down.reg.degs <- sum(
    results.df$logfc < -fc.limit & results.df$padj < padj.limit,
    na.rm = TRUE
  )

  total.up.reg.degs <- sum(
    results.df$logfc > fc.limit & results.df$padj < padj.limit,
    na.rm = TRUE
  )

  cat(
    "**DEGs at |log2FC| > ",
    fc.limit,
    " and padj < ",
    padj.limit,
    ":** ",
    total.up.reg.degs,
    " up, ",
    total.down.reg.degs,
    " down\n\n",
    sep = ""
  )

  custom.labels <- comp.params$custom.gene.labels %||% NULL

  if (isTRUE(comp.params$label.top.genes %||% FALSE)) {
    custom.labels <- get_top_gene_labels(
      results.df = results.df,
      fc.limit = fc.limit,
      padj.limit = padj.limit,
      n = comp.params$n.labels %||% 10
    )
  }

  volcano.plot <- make_de_volcano(
    results.df = results.df,
    comp.params = comp.params,
    custom.labels = custom.labels,
    remove.all.gene.labels = comp.params$remove.all.gene.labels %||% FALSE,
    label.size = comp.params$label.size %||% 8
  )

  print(volcano.plot)

  volcano.plot.nolabel <- NULL

  if (
    isTRUE(comp.params$make.nolabel.volcano %||% FALSE) ||
    isTRUE(comp.params$export.volcano.nolabel %||% FALSE)
  ) {
    volcano.plot.nolabel <- make_de_volcano(
      results.df = results.df,
      comp.params = comp.params,
      custom.labels = NULL,
      remove.all.gene.labels = TRUE,
      label.size = comp.params$nolabel.label.size %||% 4
    )
  }
  
  # Create a directory for the volcano plots
  if (
    isTRUE(comp.params$export.volcano %||% FALSE) ||
    isTRUE(comp.params$export.volcano.nolabel %||% FALSE)
  ) {
    volcano.dir <- file.path(params$de.folder, "volcano")
    
    if(!dir.exists(volcano.dir)){
      dir.create(volcano.dir, showWarnings = FALSE, recursive = TRUE)
      print(paste0("Created volcano directory at: ", volcano.dir))
    }
    
  }

  if (isTRUE(comp.params$export.volcano %||% FALSE)) {
    ggsave(
      filename = file.path(
        params$de.folder,
        "volcano",
        paste0(comp.params$contrast.name, "_volcano_plot.png")
      ),
      plot = volcano.plot,
      width = comp.params$volcano.width %||% 14,
      height = comp.params$volcano.height %||% 10
    )
    
    print(paste0("Volcano plot saved to: ", 
                 file.path(params$de.folder, "volcano",
                           paste0(comp.params$contrast.name, 
                                  "_volcano_plot.png")))
          )
    
  }

  if (isTRUE(comp.params$export.volcano.nolabel %||% FALSE)) {
    ggsave(
      filename = file.path(
        params$de.folder,
        "volcano",
        paste0(comp.params$contrast.name, "_volcano_plot_nolabel.png")
      ),
      plot = volcano.plot.nolabel,
      width = comp.params$volcano.width %||% 14,
      height = comp.params$volcano.height %||% 10
    )
  }

  gsea.preranked.df <- gsea_preranked_list(
    contrast.field = comp.params$gsea.contrast.field %||% comp.params$region.col,
    contrast.levels = comp.params$gsea.contrast.levels %||% comp.params$regions,
    annotation = lmm.input$subset.annotation,
    log.counts = lmm.input$subset.object@assayData$log_q
  )

  if (isTRUE(comp.params$export.gsea.input %||% FALSE)) {
    gsea.dir <- file.path(params$data.folder, "gsea", "input_lists")
    dir.create(gsea.dir, showWarnings = FALSE, recursive = TRUE)

    write.csv(
      gsea.preranked.df,
      file = file.path(
        gsea.dir,
        paste0(comp.params$contrast.name, "_gsea_preranked_input.csv")
      ),
      row.names = FALSE
    )
  }

  invisible(
    list(
      contrast.name = comp.params$contrast.name,
      summary.table = summary.table,
      results = results.df,
      gsea.preranked = gsea.preranked.df
    )
  )
}
```

## Comparison Parameters

```{r de-comparison-config}
#| warning: false
#| message: false


# Parameters shared across all comparisons
common.comparison.parameters <- list(
  
  # Volcano Parameters
  fc.limit = 1,
  padj.limit = 0.05,
  n.labels = 10,

  write.results = TRUE,
  overwrite.results = FALSE,

  title.size = 22,
  legend.coordinates = c(.99, .1),
  x.lab = "Log2 Fold Change",
  y.lab = "-Log10 adjusted p-value",
  dotted.line.color = "black",
  alpha = 1,
  nonDE.color = "gray60",
  upDE.color = "red",
  downDE.color = "blue",
  label.size = 8,
  nolabel.label.size = 4,
  label.color = "black",
  axis.tick.label.size = 20,
  legend.text.size = 20,
  axis.title.size = 20,

  volcano.width = 14,
  volcano.height = 10, 
  
  label.top.genes = FALSE,
  make.nolabel.volcano = TRUE,

  export.volcano = TRUE,
  export.volcano.nolabel = TRUE,
  export.gsea.input = TRUE
)

# Individual Comparisons with unique parameters
de.comparisons <- list(
  
  # Add unique paramaters for each comparison
  modifyList(
    common.comparison.parameters,
    list(
      contrast.name = "DKD_Tubule_panck_pos_vs_neg",
      heading = "DKD Tubule PanCK Pos vs. Neg",

      subset.group.1 = list(
        class = c("DKD"),
        region = c("tubule"),
        segment = c("PanCK_pos")
      ),

      subset.group.2 = list(
        class = c("DKD"),
        region = c("tubule"),
        segment = c("PanCK_neg")
      ),

      # DE function
      analysis.type = "Within Groups",
      region.col = "segment",
      regions = c("PanCK_pos", "PanCK_neg"),
      group.col = "region",
      groups = c("tubule"),
      legend.title = "Expression in PanCK+",

      # Volcano Plot
      volcano.title = "DKD Tubule - PanCK Pos vs. Neg",
      label.top.genes = FALSE, 
      
      # GSEA
      gsea.contrast.field = "segment",
      gsea.contrast.levels = c("PanCK_pos", "PanCK_neg")

    )
  ), 
  # Add unique paramaters for each comparison
  modifyList(
    common.comparison.parameters,
    list(
      contrast.name = "Glomerulus_DKD_vs_Normal",
      heading = "Glomerulus DKD vs Normal",

      subset.group.1 = list(
        class = c("DKD"), 
        region = c("glomerulus")
      ),

      subset.group.2 = list(
        class = c("normal"), 
        region = c("glomerulus")
      ),

      # DE function
      analysis.type = "Between Groups",
      region.col = "region",
      regions = c("glomerulus"),
      group.col = "class",
      groups = c("DKD", "normal"),
      legend.title = "Expression in DKD",

      # Volcano Plot
      volcano.title = "Glomerulus DKD vs. Normal",
      label.top.genes = FALSE, 
      
      # GSEA
      gsea.contrast.field = "class",
      gsea.contrast.levels = c("DKD", "normal")

    )
  ),
  
  modifyList(
    common.comparison.parameters,
    list(
      contrast.name = "DKD Glomerulus_vs_Tubule",
      heading = "DKD Glomerulus vs Tubule",

      subset.group.1 = list(
        class = c("DKD"), 
        region = c("glomerulus", 
                   "tubule")
      ),

      subset.group.2 = list(
        class = c("DKD"), 
        region = c("glomerulus", 
                   "tubule")
      ),

      # DE function
      analysis.type = "Within Groups",
      region.col = "region",
      regions = c("glomerulus", "tubule"),
      group.col = "class",
      groups = c("DKD"),
      legend.title = "Expression in Glomerulus",

      # Volcano Plot
      volcano.title = "DKD Glomerulus vs. Tubule",
      label.top.genes = FALSE, 
      
      # GSEA
      gsea.contrast.field = "region",
      gsea.contrast.levels = c("glomerulus", "tubule")

    )
  )
  
  

)

```

#### Run Comparisons

```{r run-de-comparisons}
#| warning: false
#| message: false
#| results: asis
#| fig-width: 14
#| fig-height: 10

de.outputs <- stats::setNames(
  vector("list", length(de.comparisons)),
  vapply(de.comparisons, function(x) x$contrast.name, character(1))
)

for (i in seq_along(de.comparisons)) {
  de.outputs[[i]] <- run_de_comparison(
    comp.params = de.comparisons[[i]],
    normalized.object = normalized.object,
    params = params
  )
}


```

### Set up the Annotation Colors

```{r Annotation Colors}

# Set up annotation colors for the heatmaps
anno.colors = list(
      segment = c(GeometricSegment = "slateblue", 
                PanCK_pos = "coral", 
                PanCK_neg = "seagreen1"), 
      region = c(glomerulus = "red", 
                 tubule = "blue"), 
      class = c(DKD = "purple", 
                normal = "yellow"), 
      pathology = c(normal = "orange", 
                    abnormal = "green", 
                    "NA" = "gray")
      )
```

### Run DE

#### DKD vs. normal

```{r}
#| warning: false
#| message: false

# Name for the contrast 
contrast.name <- "Original_DKD_Tubule_PanCK_Pos_vs_Neg"

subset.group.1 <- list(class = c("DKD"))

subset.group.2 <- list(class = c("normal"))

lmm.input <- subset_object_for_lmm(object = normalized.object, 
                      subset.group.1 = subset.group.1, 
                      subset.group.2 = subset.group.2)


# Create summary table of group counts
summary.table.df <- pData(lmm.input$subset.object) %>%  
  select(c(names(subset.group.1)))

summary.table <- table(summary.table.df)

print("Sample Numbers per Annotation Group")
print("-----------------------------------")
print(summary.table)

# Check if the DE results have already been generated
DE.result.file <- grepl(contrast.name, DE.results.files)
if(sum(DE.result.file) == 1) {
  
  # Load the previously generated DE results
  results.df <- as.data.frame(read.csv(paste0(params$data.folder, params$results.folder, contrast.name, "_de_results.csv")))
                              
} else {
  # Use the function from the DSPWorkflow package

  # Within slide analysis
  # Listed contrasts are condition, reference 
  results.list <- diffExpr(object = lmm.input$subset.object, 
                        analysis.type = "Between Groups", 
                        region.col = "fixed_value_column", 
                        regions = c("value"), 
                        group.col = "class", 
                        groups = c("DKD", "normal"), 
                        n.cores = parallel::detectCores())
  
  # Create the results df
  results.df <- results.list$results
  
  # Adjust column names
  logfc.column <- colnames(results.df[grepl("logFC",colnames(results.df))])
  results.df$logfc <- results.df[[logfc.column]]
  
  pval.column <- colnames(results.df[grepl("_pval",colnames(results.df))])
  results.df$pval <- results.df[[pval.column]]
  
  adj.pval.column <- colnames(results.df[grepl("adjpval",colnames(results.df))])
  results.df$padj <- results.df[[adj.pval.column]]
  
  results.df$gene <- results.df$Gene
  
  # Keep only the necessary columns
  results.df <- results.df %>% select(c("gene", 
                                        "logfc", 
                                        "pval", 
                                        "padj"))
  
  write.results <- FALSE
  if(write.results == TRUE){
    write.csv(results.df, paste0(params$data.folder, params$results.folder, contrast.name, "_de_results.csv"))
    
    saveRDS(results.df, paste0(params$data.folder, 
                              params$results.folder, 
                              contrast.name, "_de_results.RDS"))
    
  }
  
  
}




# Export the results
export.deg.list <- FALSE
if(export.deg.list == TRUE){ 
  write.csv(results.df, 
            file = paste0(results.folder, run.folder, "DE/DEG_lists/", contrast.name, "_deg_list.csv"), 
            row.names = FALSE)  
}

# Make the volcano plot
volcano.output <- make_volcano(lmm.results = results.df, 
                               title = contrast.name, 
                               legend.title = "Expression in DKD", 
                               x.axis.title = "logFC", 
                               fc.limit = 1.5, 
                               pos.label.limit = 1.5, 
                               neg.label.limit = -1.5)

# Export the volcano
export.volcano <- FALSE
if(export.volcano == TRUE){ 
  ggsave(filename = paste0(results.folder, run.folder, "DE/volcano/", contrast.name, "_volcano_plot.png"), 
         width = 14, 
         height = 10)
}

print(volcano.output$volcano.plot)

# Subset the annotation just for the heatmap annotations of interest
#annotation.heatmap <- lmm.input$annotation %>% 
#  select(names(anno.colors)) %>% 
#  arrange(Treatment_group)
#
#heatmap.plot <- make_heatmap(
#  normalized.log.counts.df = lmm.input$log.counts, 
#  de.results = results.df, 
#  top.degs = TRUE, 
#  annotation.column = annotation.heatmap,
#  annotation.row = NULL, 
#  anno.colors = anno.colors, 
#  cluster.rows = TRUE, 
#  cluster.columns = TRUE, 
#  main.title = contrast.name, 
#  row.gaps = NULL, 
#  column.gaps = NULL, 
#  show.colnames = FALSE, 
#  show.rownames = TRUE)
#
#print(heatmap.plot)

#export.heatmap <- FALSE
#if(export.heatmap == TRUE){
#  
#  ggsave(heatmap.plot, 
#         filename = paste0(results.folder, run.folder, "DE/heatmap/", contrast.name, "_heatmap_plot#.png"), 
#         width = 14, 
#         height = 10)
#  
#}
#
## GSEA Preranked List
#
#gsea.preranked.df <- gsea_preranked_list(contrast.field = contrast.field, 
#                                         contrast.levels = contrast.groups, 
#                                         annotation = lmm.input$annotation,
#                                         log.counts = lmm.input$log.counts)
#
#export.gsea.input <- FALSE
#if(export.gsea.input == TRUE){
#  
#  write.csv(GSEA.preanked.df, file = paste0(results.folder, run.folder, "GSEA/", contrast.name, #"_gsea_preranked_input.csv"), row.names = FALSE)
#  
#}

```

#### DKD - Glomerulus vs. Tubule

```{r}
#| warning: false
#| message: false

# Name for the contrast 
contrast.name <- "DKD_glomerulus_vs_tubule"

subset.group.1 <- list(class = c("DKD"), 
                       region = c("glomerulus"))

subset.group.2 <- list(class = c("DKD"), 
                       region = c("tubule"))

lmm.input <- subset_object_for_lmm(object = normalized.object, 
                      subset.group.1 = subset.group.1, 
                      subset.group.2 = subset.group.2)


# Create summary table of group counts
summary.table.df <- pData(lmm.input$subset.object) %>%  
  select(c(names(subset.group.1)))

summary.table <- table(summary.table.df)

print("Sample Numbers per Annotation Group")
print("-----------------------------------")
print(summary.table)

# Check if the DE results have already been generated
DE.result.file <- grepl(contrast.name, DE.results.files)
if(sum(DE.result.file) == 1) {
  
  # Load the previously generated DE results
  results.df <- as.data.frame(read.csv(paste0(params$data.folder, params$results.folder, contrast.name, "_de_results.csv")))
                              
} else {
  # Use the function from the DSPWorkflow package

  # Within slide analysis
  # Listed contrasts are condition, reference 
  results.list <- diffExpr(object = lmm.input$subset.object, 
                        analysis.type = "Within Groups", 
                        region.col = "region", 
                        regions = c("glomerulus", "tubule"), 
                        group.col = "class", 
                        groups = c("DKD"), 
                        n.cores = parallel::detectCores())
  
  # Create the results df
  results.df <- results.list$results
  
  write.results <- FALSE
  if(write.results == TRUE){
    write.csv(results.df, paste0(params$data.folder, params$results.folder, contrast.name, "_de_results.csv"))
    
    saveRDS(results.df, paste0(params$data.folder, 
                              params$results.folder, 
                              contrast.name, "_de_results.RDS"))
    
  }
  
  
}

# Adjust column names
logfc.column <- colnames(results.df[grepl("logFC",colnames(results.df))])
results.df$logfc <- results.df[[logfc.column]]

pval.column <- colnames(results.df[grepl("_pval",colnames(results.df))])
results.df$pval <- results.df[[pval.column]]

adj.pval.column <- colnames(results.df[grepl("adjpval",colnames(results.df))])
results.df$padj <- results.df[[adj.pval.column]]

results.df$gene <- results.df$Gene

# Keep only the necessary columns
results.df <- results.df %>% select(c("gene", 
                                      "logfc", 
                                      "pval", 
                                      "padj"))


# Export the results
export.deg.list <- FALSE
if(export.deg.list == TRUE){ 
  write.csv(results.df, 
            file = paste0(results.folder, run.folder, "DE/DEG_lists/", contrast.name, "_deg_list.csv"), 
            row.names = FALSE)  
}

# Make the volcano plot
volcano.output <- make_volcano(lmm.results = results.df, 
                               title = contrast.name, 
                               legend.title = "Expression in glomerulus", 
                               x.axis.title = "logFC", 
                               fc.limit = 1, 
                               pos.label.limit = 1.5, 
                               neg.label.limit = -1.5)

# Export the volcano
export.volcano <- FALSE
if(export.volcano == TRUE){ 
  ggsave(filename = paste0(results.folder, run.folder, "DE/volcano/", contrast.name, "_volcano_plot.png"), 
         width = 14, 
         height = 10)
}

print(volcano.output$volcano.plot)

# Subset the annotation just for the heatmap annotations of interest
#annotation.heatmap <- lmm.input$annotation %>% 
#  select(names(anno.colors)) %>% 
#  arrange(Treatment_group)
#
#heatmap.plot <- make_heatmap(
#  normalized.log.counts.df = lmm.input$log.counts, 
#  de.results = results.df, 
#  top.degs = TRUE, 
#  annotation.column = annotation.heatmap,
#  annotation.row = NULL, 
#  anno.colors = anno.colors, 
#  cluster.rows = TRUE, 
#  cluster.columns = TRUE, 
#  main.title = contrast.name, 
#  row.gaps = NULL, 
#  column.gaps = NULL, 
#  show.colnames = FALSE, 
#  show.rownames = TRUE)
#
#print(heatmap.plot)

#export.heatmap <- FALSE
#if(export.heatmap == TRUE){
#  
#  ggsave(heatmap.plot, 
#         filename = paste0(results.folder, run.folder, "DE/heatmap/", contrast.name, "_heatmap_plot#.png"), 
#         width = 14, 
#         height = 10)
#  
#}
#
## GSEA Preranked List
#
#gsea.preranked.df <- gsea_preranked_list(contrast.field = contrast.field, 
#                                         contrast.levels = contrast.groups, 
#                                         annotation = lmm.input$annotation,
#                                         log.counts = lmm.input$log.counts)
#
#export.gsea.input <- FALSE
#if(export.gsea.input == TRUE){
#  
#  write.csv(GSEA.preanked.df, file = paste0(results.folder, run.folder, "GSEA/", contrast.name, #"_gsea_preranked_input.csv"), row.names = FALSE)
#  
#}


```

#### DKD - Tubule - PanCK Pos vs. Neg

```{r}

#| warning: false
#| message: false

# Name for the contrast 
contrast.name <- "DKD_glomerulus_vs_tubule"

subset.group.1 <- list(class = c("DKD"), 
                       region = c("tubule"), 
                       segment = c("PanCK_pos"))

subset.group.2 <- list(class = c("DKD"), 
                       region = c("tubule"), 
                       segment = c("PanCK_neg"))

lmm.input <- subset_object_for_lmm(object = normalized.object, 
                      subset.group.1 = subset.group.1, 
                      subset.group.2 = subset.group.2)


# Create summary table of group counts
summary.table.df <- pData(lmm.input$subset.object) %>%  
  select(c(names(subset.group.1)))

summary.table <- table(summary.table.df)

print("Sample Numbers per Annotation Group")
print("-----------------------------------")
print(summary.table)

# Check if the DE results have already been generated
DE.result.file <- grepl(contrast.name, DE.results.files)
if(sum(DE.result.file) == 1) {
  
  # Load the previously generated DE results
  results.df <- as.data.frame(read.csv(paste0(params$data.folder, params$results.folder, contrast.name, "_de_results.csv")))
                              
} else {
  # Use the function from the DSPWorkflow package

  # Within slide analysis
  # Listed contrasts are condition, reference 
  results.list <- diffExpr(object = lmm.input$subset.object, 
                        analysis.type = "Within Groups", 
                        region.col = "segment", 
                        regions = c("PanCK_pos", "PanCK_neg"), 
                        group.col = "class", 
                        groups = c("DKD"), 
                        n.cores = parallel::detectCores())
  
  # Create the results df
  results.df <- results.list$results
  
  write.results <- FALSE
  if(write.results == TRUE){
    write.csv(results.df, paste0(params$data.folder, params$results.folder, contrast.name, "_de_results.csv"))
    
    saveRDS(results.df, paste0(params$data.folder, 
                              params$results.folder, 
                              contrast.name, "_de_results.RDS"))
    
  }
  
  
}

# Adjust column names
logfc.column <- colnames(results.df[grepl("logFC",colnames(results.df))])
results.df$logfc <- results.df[[logfc.column]]

pval.column <- colnames(results.df[grepl("_pval",colnames(results.df))])
results.df$pval <- results.df[[pval.column]]

adj.pval.column <- colnames(results.df[grepl("adjpval",colnames(results.df))])
results.df$padj <- results.df[[adj.pval.column]]

results.df$gene <- results.df$Gene

# Keep only the necessary columns
results.df <- results.df %>% select(c("gene", 
                                      "logfc", 
                                      "pval", 
                                      "padj"))


# Export the results
export.deg.list <- FALSE
if(export.deg.list == TRUE){ 
  write.csv(results.df, 
            file = paste0(results.folder, run.folder, "DE/DEG_lists/", contrast.name, "_deg_list.csv"), 
            row.names = FALSE)  
}

# Make the volcano plot
volcano.output <- make_volcano(lmm.results = results.df, 
                               title = contrast.name, 
                               legend.title = "Expression in PanCK Positive", 
                               x.axis.title = "logFC", 
                               fc.limit = 1, 
                               pos.label.limit = 1.5, 
                               neg.label.limit = -1.5)

# Export the volcano
export.volcano <- FALSE
if(export.volcano == TRUE){ 
  ggsave(filename = paste0(results.folder, run.folder, "DE/volcano/", contrast.name, "_volcano_plot.png"), 
         width = 14, 
         height = 10)
}

print(volcano.output$volcano.plot)

# Subset the annotation just for the heatmap annotations of interest
#annotation.heatmap <- lmm.input$annotation %>% 
#  select(names(anno.colors)) %>% 
#  arrange(Treatment_group)
#
#heatmap.plot <- make_heatmap(
#  normalized.log.counts.df = lmm.input$log.counts, 
#  de.results = results.df, 
#  top.degs = TRUE, 
#  annotation.column = annotation.heatmap,
#  annotation.row = NULL, 
#  anno.colors = anno.colors, 
#  cluster.rows = TRUE, 
#  cluster.columns = TRUE, 
#  main.title = contrast.name, 
#  row.gaps = NULL, 
#  column.gaps = NULL, 
#  show.colnames = FALSE, 
#  show.rownames = TRUE)
#
#print(heatmap.plot)

#export.heatmap <- FALSE
#if(export.heatmap == TRUE){
#  
#  ggsave(heatmap.plot, 
#         filename = paste0(results.folder, run.folder, "DE/heatmap/", contrast.name, "_heatmap_plot#.png"), 
#         width = 14, 
#         height = 10)
#  
#}
#
## GSEA Preranked List
#
#gsea.preranked.df <- gsea_preranked_list(contrast.field = contrast.field, 
#                                         contrast.levels = contrast.groups, 
#                                         annotation = lmm.input$annotation,
#                                         log.counts = lmm.input$log.counts)
#
#export.gsea.input <- FALSE
#if(export.gsea.input == TRUE){
#  
#  write.csv(GSEA.preanked.df, file = paste0(results.folder, run.folder, "GSEA/", contrast.name, #"_gsea_preranked_input.csv"), row.names = FALSE)
#  
#}


```

### Venn Diagrams for log FC cutoff of 1

```{r Setup Gene Lists log FC 1}
#| warning: false
#| message: false

# Load all DE results
results.files <- list.files(paste0(results.folder, run.folder))

up.results.list <- list()
down.results.list <- list()

logfc.cutoff <- 1

results.files <- results.files[grepl(".csv", results.files)]


for(file in results.files){
  
  # Load the results for a DE contrast
  de.results <- as.data.frame(read.csv(paste0(results.folder, run.folder, file)))
  
  # Filter for DEGs for up and down regulated
  logfc.column <- colnames(de.results)[grep("logFC", colnames(de.results))]
  adj.pval.column <- colnames(de.results)[grep("adjpval", colnames(de.results))]
  
  up.degs.data <- de.results %>% 
    filter(.data[[logfc.column]] > logfc.cutoff) %>% 
    filter(.data[[adj.pval.column]] < 0.05)
  
  down.degs.data <- de.results %>% 
  filter(.data[[logfc.column]] < -(logfc.cutoff)) %>% 
  filter(.data[[adj.pval.column]] < 0.05)
  
  # Gather the DE gene names and store
  up.degs <- up.degs.data$Gene
  down.degs <- down.degs.data$Gene
  
  contrast.name <- gsub("_de.results.csv", "", file)
  contrast.name <- gsub("_", " vs. ", contrast.name)
  
  if(length(up.degs) > 0){
    
    up.results.list[[contrast.name]] <- up.degs
    
  } else {
    
    # For group with no DEGs
    up.results.list[[contrast.name]] <- "no DEGs"
    
  }
  
  if(length(down.degs) > 0){
    
    down.results.list[[contrast.name]] <- down.degs
    
  } else {
    
    # For group with no DEGs
    down.results.list[[contrast.name]] <- "no DEGs"
    
  }
  
}
```

#### Up-regulated Venn Diagram

```{r Create venn diagram for up degs log FC 1}
#| warning: false
#| message: false


venn.up.degs <- venn.diagram(up.results.list, 
                            main = "Up-regulated DEGs each DE Contrast",
                            category.names = c(names(up.results.list)[1],
                                               names(up.results.list)[2], 
                                               names(up.results.list)[3], 
                                               names(up.results.list)[4]), 
                            fill = c("maroon", 
                                     "#E69F00", 
                                     "#56B4E9", 
                                     "#009E73"), 
                            filename = NULL)


grid.draw(venn.up.degs)
```

```         
```

#### Boxplots

```{r Deconvolution Boxplots}
#| warning: false
#| message: false

tumor.vessel.comparison <- list(
  c("tumor", "vessel"))

tumor.immune.comparison <- list(
  c("tumor", "immune"))

vessel.immune.comparison <- list(
  c("vessel", "immune"))

skin.box.plot <- ggplot(data = skin.cell.melt, aes(x = region, 
                                                   y = value, 
                                                   fill = segment)) + 
  geom_boxplot(color = "black") + 
  facet_wrap(~cell.type, scales = "free_x") + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  ) + 
  labs(y = "Percentage of Total") + 
  stat_compare_means(comparisons = tumor.vessel.comparison, 
                     label = "p.signif", 
                     label.y = 20) + 
  stat_compare_means(comparisons = tumor.immune.comparison, 
                     label = "p.signif", 
                     label.y = 40) + 
  stat_compare_means(comparisons = vessel.immune.comparison, 
                     label = "p.signif", 
                     label.y = 30)

immune.box.plot <- ggplot(data = immune.cell.melt, aes(x = region, 
                                                   y = value, 
                                                   fill = segment)) + 
  geom_boxplot(color = "black") + 
  facet_wrap(~cell.type, scales = "free_x") + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  ) + 
  labs(y = "Percentage of Total") + 
  stat_compare_means(comparisons = tumor.vessel.comparison, 
                     label = "p.signif", 
                     label.y = 20) + 
  stat_compare_means(comparisons = tumor.immune.comparison, 
                     label = "p.signif", 
                     label.y = 40) + 
  stat_compare_means(comparisons = vessel.immune.comparison, 
                     label = "p.signif", 
                     label.y = 30)

bahman.box.plot <- ggplot(data = bahman.cell.melt, aes(x = region, 
                                                   y = value, 
                                                   fill = segment)) + 
  geom_boxplot(color = "black") + 
  facet_wrap(~cell.type, scales = "free_x") + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  ) + 
  labs(y = "Percentage of Total") + 
  stat_compare_means(comparisons = tumor.vessel.comparison, 
                     label = "p.signif", 
                     label.y = 20) + 
  stat_compare_means(comparisons = tumor.immune.comparison, 
                     label = "p.signif", 
                     label.y = 40) + 
  stat_compare_means(comparisons = vessel.immune.comparison, 
                     label = "p.signif", 
                     label.y = 30)

# Export Boxplots
export.decon.boxplots <- global.export.decon
if(export.decon.boxplots == TRUE){
  
  ggsave(paste0(results.folder, 
                run.folder, 
                "spatial_decon/spatial_decon_skin_ref_region_boxplot.png"), 
         skin.box.plot, 
         height = 10, 
         width = 14)
  
  ggsave(paste0(results.folder, 
                run.folder, 
                "spatial_decon/spatial_decon_immune_ref_region_boxplot.png"), 
         immune.box.plot, 
         height = 10, 
         width = 14)
  
  ggsave(paste0(results.folder, 
                run.folder, 
                "spatial_decon/spatial_decon_bahman_ref_region_boxplot.png"), 
         bahman.box.plot, 
         height = 10, 
         width = 14)
  
}

```

## StandR Analysis

### Inputs

```{r}

# Load the StandR RUV object
standr.ruv.object <- readRDS(file = paste0(params$data.folder, "StandR/standr_ruv_object.RDS"))

# Convert to DGE for limma compatibility
ruv.dge.object <- SE2DGEList(standr.ruv.object)



```

### Set up for Limma

```{r}

# The biological field to compare


# Set up the design formula
design <- model.matrix(~0 + combined_coll_annotations + ruv_W1 + ruv_W2 , data = colData(standr.ruv.object))

colnames(design)
colnames(design) <- gsub("^combined_coll_annotations","",colnames(design))
colnames(design)

contr.matrix <- makeContrasts(
   Abnormal_v_Normal = glomerulus_abnormal_GeometricSegment - glomerulus_normal_GeometricSegment,
   levels = colnames(design))

# Filter genes by expression
keep <- filterByExpr(ruv.dge.object, design)

# Genes removed
table(keep)
rownames(ruv.dge.object)[!keep]

# Remove low expression genes
ruv.dge.object.gene.filter <- ruv.dge.object[keep, ]

# Extimate Biological Coeff of Variation
ruv.dge.object.gene.filter <- estimateDisp(ruv.dge.object.gene.filter, 
                                           design = design, 
                                           robust = TRUE)


```

#### BCV Plot

```{r}


bcv.df <- data.frame(
  'BCV' = sqrt(ruv.dge.object.gene.filter$tagwise.dispersion),
  'AveLogCPM' = ruv.dge.object.gene.filter$AveLogCPM,
  'gene_id' = rownames(ruv.dge.object.gene.filter)
)

bcv.df$highBCV <- bcv.df$BCV > 0.8

if (!is.null(ruv.dge.object.gene.filter$trended.dispersion)) {
  bcv.df$trendedBCV <- sqrt(ruv.dge.object.gene.filter$trended.dispersion)
}

bcv.plot <- ggplot(bcv.df, aes(x = AveLogCPM, y = BCV)) +
  geom_point(
    aes(color = highBCV),
    alpha = 0.6,
    size = 1.5) +
  geom_text_repel(
    data = subset(bcv.df, highBCV),
    aes(label = gene_id),
    size = 3,
    max.overlaps = Inf) +
  scale_color_manual(
    values = c("FALSE" = "black", "TRUE" = "red"),
    labels = c("Other genes", "BCV > 0.8"),
    name = NULL) +
  coord_cartesian(ylim = c(0, 1.3)) +
  labs(
    x = "Average log CPM",
    y = "Biological coefficient of variation",
    title = "BCV plot") +
  theme_classic() +
  theme(
    legend.position = c(0.02, 0.98),
    legend.justification = c(0, 1))

# Add trended BCV line, if available
if ("trendedBCV" %in% colnames(bcv.df)) {
  bcv.plot <- bcv.plot +
    geom_line(
      data = bcv.df[order(bcv.df$AveLogCPM), ],
      aes(y = trendedBCV),
      color = "blue",
      linewidth = 1
    )
}

# Add common BCV line, if available
if (!is.null(ruv.dge.object.gene.filter$common.dispersion)) {
  bcv.plot <- bcv.plot +
    geom_hline(
      yintercept = sqrt(ruv.dge.object.gene.filter$common.dispersion),
      color = "blue",
      linetype = "dashed",
      linewidth = 0.8
    )
}



```

#### Run Limma Voom

```{r}


object.metadata <- as.data.frame(colData(standr.ruv.object))

# Define the field which represents biological samples
# Slide should be used in the case of 1 subject per slide, otherwise a patient or animal ID should be used
block <- object.metadata$slide_name

# Estimate mean-variance trend, first pass
v0 <- voom(ruv.dge.object.gene.filter, 
           design = design, 
           plot = TRUE)

# Estimate the within block correlation
corfit1 <- duplicateCorrelation(
  v0,
  design = design,
  block = block
)

# Correlation for first pass
corfit1$consensus.correlation

# Estimate mean variance trend now with first block correlation
v <- voom(
  ruv.dge.object.gene.filter,
  design = design,
  block = block,
  correlation = corfit1$consensus.correlation,
  plot = TRUE
)

# Estimate the block correlation again with the second pass mean variance trend
corfit2 <- duplicateCorrelation(
  v,
  design = design,
  block = block
)

# Correlation for second pass
corfit2$consensus.correlation

# Linear model for each gene
fit <- lmFit(
  v,
  design = design,
  block = block,
  correlation = corfit2$consensus.correlation
)


fit_contrast <- contrasts.fit(
  fit,
  contrasts = contr.matrix
)

# Final 
efit <- eBayes(
  fit_contrast,
  robust = TRUE
)


results <- topTable(efit, 
                    number = Inf)

results.simple <- results %>% 
  select(mean_expr, logFC, adj.P.Val)


results.simple.degs <- results.simple %>% 
  filter(adj.P.Val < 0.05) %>% 
  filter(logFC < -1 | logFC > 1)
```

## Compare DEGS Default vs. StandR

```{r}


```
